---
title: "Project-5 Markdown"
author: "Vishal Balaji"
date: 2020-04-29
output: 
  html_notebook:
    theme: readable
    toc: true
    toc_float: true
    highlight: kate
  pdf_document: default
---

```{r setup, include = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("F:/Google Drive/Classes/2020 Spring/data analysis")
imageDirectory <- "images"
```


## Preliminaries

```{r libs, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r functions, include = FALSE}
saveInImageDirectory <- function(filename){
  imageFile <- file.path(imageDirectory, filename)
  ggsave(imageFile, width = 30, height = 20, units = "cm", dpi = 300)
}
```


## Data
This code below imports the data with the proper column data type.
```{r import}
tuitionData <- read_csv("Data/Tuition.csv", 
                        col_types = cols(
                          Name = col_character(), 
                          Location = col_character(), 
                          `In-State Tuition` = col_double(), 
                          `Out-of-State Tuition` = col_double(), 
                          Type = col_factor()
                        ))
```
The data is then mutated to include three new columns:    

* `Classification` is of type factor containing three levels:
  + College
  + University
  + Community College
  
  This is done by comparing the insitiution name and extracting the classification which is then converted into a factor. All insitutions that do not have any of the classification in their `name` will have an "NA" factor. This will be changed into a factor of "College"

* `State` is of type factor containing 45 levels, each level represents a state
* `Location_print` is of type character which contains the `Location` with proper capitalization

```{r data-formatting}
tuitionData <- tuitionData %>% 
  mutate(
    Classification = parse_factor(
      str_extract(tuitionData$Name, "College|University|Community College")
      ),
    State = parse_factor(str_extract(tuitionData$Location, '\\S+$')),
    Location_print = 
      str_c(str_to_title(sub("(,).*", "", tuitionData$Location)), 
            str_extract(tuitionData$Location, '\\S+$'), 
            sep = ", "
            )
  )
tuitionData$Classification <- fct_explicit_na(tuitionData$Classification, "College")

```


## Colleges in Top 10

This table shows the colleges that are in the list of top 10 instituitons with the highest in-state tuition

```{r top-10}
tuitionData %>% 
  top_n(10, `In-State Tuition`) %>%
  filter(Classification == "College") %>%
  select(Name, Type, `In-State Tuition`)
```


## In-State vs Out-Of-State 

This code shows the In-State vs Out-Of-State tuition difference for private colleges

```{r inVsOutDiff}
tuitionDifference <- tuitionData %>% 
  mutate(
    Tuition_Diff = `In-State Tuition` - `Out-of-State Tuition`
  ) %>%
  filter(Type == "Private") %>%
  select(Name, Type, `Tuition_Diff`)
tuitionDifference
```


## Lower tuition than Millsaps

This section finds the private colleges that have lower tuition than millsaps

```{r lowerThanMillsaps}
millsapsTuition <- tuitionData %>% filter(str_detect(tuitionData$Name, "Millsaps"))

lowerThanMillsaps <- tuitionData %>%
  filter(Type == "Private" & Classification == "College") %>%
  filter(`In-State Tuition` < millsapsTuition$`In-State Tuition`, `Out-of-State Tuition` < millsapsTuition$`Out-of-State Tuition`)
lowerThanMillsaps
```


## Mean In-state tuition by state

This section finds the average (mean) in-state tuition for private colleges by state

```{r meanInStateTuition}
inStateTuition_byState <- tuitionData %>% 
  group_by(State) %>% 
  summarise(
    `In-State Tuition` = mean(`In-State Tuition`, na.rm = TRUE)
    ) %>% 
  arrange(desc(`In-State Tuition`))
inStateTuition_byState
```


### Mississippi's rank

This section finds the rank of Mississippi compared to other states for average in-state tuition. (1 - highest, nth - lowest)

```{r MSRank}
which(grepl("MS", inStateTuition_byState$State))
```


## Private colleges in Mississippi

```{r MS-priv-coll}
privateCollegesInMS <- tuitionData %>% 
  filter(Type == "Private", 
         Classification == "College",
         State == "MS"
         ) %>%
  arrange(desc(`In-State Tuition`))
privateCollegesInMS
```

### Millsaps's rank in Mississippi

This section finds the rank of Millsaps College compared to other private colleges in Mississippi for in-state tuition. (1 - highest, nth - lowest)

```{r millRank}
which(grepl("Millsaps", privateCollegesInMS$Name))
```


## Bar Graphs

### By state
The bar graphs shows the in-state tuition of private colleges for Alabama, Arkansas, Florida, Georgia, Louisiana, Mississippi, and Tennessee.

The function below makes the code tidier and resuable for any state.
```{r barGraph-singleState-func}
privateCollegeBarGraph <- function(state_name){
  tuitionData %>% 
    filter(Type == "Private", 
           Classification == "College",
           State == state_name
    ) %>% 
    ggplot(aes(x = Name, y = `In-State Tuition`)) + 
    theme_minimal() + 
    theme(plot.title = element_text(color = "grey25", size = 20, face = "bold")) +  
    geom_bar(stat = "identity", fill = "darkcyan", width = 0.7) +
    ggtitle(
      str_c(c("In-state Tution for Private Colleges in", state_name), collapse = " ")
    ) + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 25))
  #saveInImageDirectory(str_c(c("In-state tuition - ", state_name, ".png"), collapse = ""))
}
```

```{r barGraph-singleState, fig.width = 8, fig.height= 4}
privateCollegeBarGraph("AL")
privateCollegeBarGraph("AR")
privateCollegeBarGraph("FL") # No Private Colleges
privateCollegeBarGraph("GA")
privateCollegeBarGraph("LA")
privateCollegeBarGraph("MS")
privateCollegeBarGraph("TN")
```

### Selective states

The bar graph below shows the average in-state tuition of private colleges for Alabama, Arkansas, Florida, Georgia, Louisiana, Mississippi, and Tennessee.

```{r barGraph-selStates, fig.width = 8, fig.height= 4}
selStates = c("AL", "AR", "FL", "GA", "LA", "MS", "TN")
inStateTuition_byState %>% 
  filter(
    State %in% selStates
  ) %>%
  ggplot(aes(x = State, y = `In-State Tuition`)) + 
  theme_minimal() + 
  theme(plot.title = element_text(color = "grey25", size = 20, face = "bold")) +  
  geom_bar(stat = "identity", fill = "coral1", width = 0.7) + 
  ggtitle("Average In-state Tution by State")
```
